---
title: "Network Based Bata Analysis"
author: "Vittoria"
date: "`r Sys.Date()`"
output: html_document
---

# Preliminary steps
load libraries
```{r}
library("GEOquery")
library("useful")
library("rgl")
library("plotly")
library("randomForest")
library("RColorBrewer")
library("glmnet") 
library("ROCR")
library("genefilter")
library("caret")
library("e1071")
library("rScudo")
library("igraph")
library("KEGGREST")
library("KEGGgraph")
library("AnnotationDbi")
library("org.Hs.eg.db")
library("pathfindR")
library("biomaRt")
```

Import data

```{r}
gse <- getGEO("GSE84046", destdir = ".", getGPL = FALSE)
gse <- gse[[1]]
ex <- exprs(gse)
```

# Initial analysis

```{r}
dim(ex)
boxplot(ex, main = "Boxplot of initial data")
```


## creating vector of labels for further analysis
labels -> plane color labels
groups_labels -> labels with group names
complete_labels -> combination of group + name of the sample
y_01_control_vs_diet -> labels 0 1 as control 
and some others


```{r}
#set labels for pca
labels = c()
group_labels = c()
complete_labels = c()
control_vs_diet_01 = c()
diet_vs_diet_01 = c()
ER_positions = c()
ER_labels = c()
for (x in 1:44) {
    # BEFORE SAMPLING -> control group
    if (gse$characteristics_ch1.2[x] == "time of sampling (before/after): before"){
        labels[x] <- "green"
        group_labels[x] <- "control"
        complete_labels[x] <- paste(colnames(ex)[x], "(", "control", ")")
        control_vs_diet_01[x] <- "0"
    } 
    # AFTER SAMPLING
    else {
        control_vs_diet_01[x] <- "1"
        if (gse$characteristics_ch1.1[x] == "protein content restricted diet: high") {
            labels[x] <- "red"
            group_labels[x] <- "ER_HP"
            complete_labels[x] <- paste(colnames(ex)[x], "(", "ER_HP", ")")
            diet_vs_diet_01 = append(diet_vs_diet_01, 0)
            ER_positions = append(ER_positions, x)
            ER_labels = append(ER_labels, "HP")
        } else {
            labels[x] <- "blue"
            group_labels[x] <- "ER_SP"
            complete_labels[x] <- paste(colnames(ex)[x], "(", "ER_SP", ")")
            diet_vs_diet_01 = append(diet_vs_diet_01, 1)
            ER_positions = append(ER_positions, x)
            ER_labels = append(ER_labels, "SP")
        }
    }
}
```


## Filtering the dataset by inspecting p-values 
keep only probes that have a significant difference in at least one of the comparison
control vs HP
control vs SP
HP vs SP

-> filtered dataset is called *ex2*

```{r}
# performing t-test, will perform pairwise comparison (3 total t-tests)
# this test is done for feature selection
# I avoid considering features with a p-value > 0.01 in one of the three t-tests

p_values_HP_NP <- c()
p_values_NP_control <- c()
p_values_HP_control <- c()

for (i in 1:32297){
  # for each mRNA expression row ex[i, 0]
  HP = c()
  NP = c()
  control = c()
  
  for (j in 1:44){
    # for each sample in column j ex[i, j]
    
    if (group_labels[j] == "control"){
      # control case
      control <- append(control, 2^ex[i, j])
    } else if (group_labels[j] == "ER_HP") {
      # HP case
      HP <- append(HP, 2^ex[i, j])
    } else {
      # NP case
      NP <- append(NP, 2^ex[i, j])
    }
  }
  
  t_NP_HP <- t.test(HP, NP)
  p_values_HP_NP <- append(p_values_HP_NP, t_NP_HP$p.value)
  t_NP_control <- t.test(NP, control)
  p_values_NP_control <- append(p_values_NP_control, t_NP_control$p.value)
  t_HP_control <- t.test(HP, control)
  p_values_HP_control <- append(p_values_HP_control, t_HP_control$p.value)
  
}

threshold <- 0.01
# create empty new matrix
ex2 = matrix(,nrow = 0, ncol = 44)
ex2_unique_probes = matrix(,nrow = 0, ncol = 44)
colnames(ex2) <- colnames(ex)
colnames(ex2_unique_probes) <- colnames(ex)

for (i in 1 : 32297){
  if (p_values_HP_control[i] < threshold ||
      p_values_HP_NP[i] < threshold ||
      p_values_NP_control[i] < threshold) {
    # interesting values
    ex2 <- rbind(ex2, ex[i, ])
    ex2_unique_probes <- rbind(ex2_unique_probes, ex[i, ])
    rownames(ex2)[dim(ex2)[1]] <- rownames(ex)[i] 
    rownames(ex2_unique_probes)[dim(ex2_unique_probes)[1]] <- rownames(ex)[i] 
  }
}
```

# PCA

### performing 2d - PCA with first two components 

```{r}
pca <- prcomp(t(ex2))
screeplot(pca)

components <- pca$x
components <- data.frame(components)
components <- cbind(components, group_labels)

fig <- plot_ly(components, x = ~PC1, y = ~PC2, color = group_labels, colors =  c('#636EFA','#EF553B','#00CC96'), type = "scatter", mode = "markers", title="PCA analysis with whole dataset")

fig

```

### performing 3d - PCA

```{r}
components <- pca$x
components <- data.frame(components)
components$PC2 <- -components$PC2
components$PC3 <- -components$PC3
components = cbind(components, group_labels)

tot_explained_variance_ratio <- summary(pca)[["importance"]]['Proportion of Variance',]
tot_explained_variance_ratio <- 100 * sum(tot_explained_variance_ratio)

tit = '3D pca plot'
fig <- plot_ly(components, x = ~PC1, y = ~PC2, z = ~PC3, color = group_labels, colors = c('#636EFA','#EF553B','#00CC96') ) %>%
  add_markers(size = 12)
fig <- fig %>%
  layout(
    title = tit,
    scene = list(bgcolor = "#e5ecf6")
)

fig
```


# Clustering

### k-means

```{r}
k <- 3
set.seed(2900)
kmeans_results = kmeans(t(ex2), k)
table(kmeans_results$cluster)
plot(kmeans_results, data=t(ex2)) + geom_text(aes(label = group_labels), hjust = 0, vjust = 0)
```

### hierarchical clustering


```{r}
dist_matrix = dist(t(ex2))
hc_results = hclust(dist_matrix, method = "ave")

k <- 3
groups = cutree(hc_results, k=k)
# plot(hc_results, hang <-1, labels = groups)
plot(hc_results, hang <-1, labels = group_labels, main = "hierarchical clustering")
# rect.hclust(hc_results, k=3, which = NULL, x = NULL, h = NULL, cluster = NULL, border = 2)
```

```{r}
dist_matrix = dist(t(ex2))
hc_results = hclust(dist_matrix, method = "complete")

k <- 3
groups = cutree(hc_results, k=k)
# plot(hc_results, hang <-1, labels = groups)
plot(hc_results, hang <-1, labels = group_labels, main = "hierarchical clustering")
# rect.hclust(hc_results, k=3, which = NULL, x = NULL, h = NULL, cluster = NULL, border = 2)
```

```{r}
dist_matrix = dist(t(ex2))
hc_results = hclust(dist_matrix, method = "single")

k <- 3
groups = cutree(hc_results, k=k)
# plot(hc_results, hang <-1, labels = groups)
plot(hc_results, hang <-1, labels = group_labels, main = "hierarchical clustering")
# rect.hclust(hc_results, k=3, which = NULL, x = NULL, h = NULL, cluster = NULL, border = 2)
```

# Supervised analysis

## Random forest

random forest error
```{r}
set.seed(2000)
rf <- randomForest(x = t(ex2), y = as.factor(group_labels), ntree=1000)
plot(rf, main = "random forest error")
```

gene rank for importance of random forest classification

```{r}
imp.temp <- abs(rf$importance)
t <- order(imp.temp, decreasing = TRUE)

plot(c(1:nrow(ex2)), imp.temp[t], log="x", cex.main = 1.5, 
     xlab = "gene rank", ylab="variable importance", cex.lab = 1.5,
     phc = 16, main = "all subset results")
```

heatmap
```{r}
gn.imp <- names(imp.temp)[t]
gn.25 <- gn.imp[1:25]
t <- is.element(rownames(ex2), gn.25)
sig.ex <- ex2[t,]

# make heatmap
hmcol <- colorRampPalette(brewer.pal(11, "RdPu"))(500)
colnames(ex2) <- group_labels
csc <- rep(hmcol[50], 30)
csc[group_labels == "control"] <- hmcol[100]
csc[group_labels == "ER_HP"] <- hmcol[300]
csc[group_labels == "ER_SP"] <- hmcol[400]
heatmap(ex2, scale="row", col=hmcol, ColSideColors = csc, main = "Heatmap for gene espression")
```

## LDA, RF and LASSO

trial: how much does it change by losing the control samples? -> enough
```{r}
#subsetting the ex2 with only ER_SP and ER_HP
ex3 <- ex2[, ER_positions]
ex3_unique_probes <- ex2_unique_probes[, ER_positions]

set.seed(2000)
rf <- randomForest(x = t(ex3), y = as.factor(ER_labels), ntree=1000)
plot(rf, main = "random forest error on ER samples")
```

check importance for every probe used in random forest training

```{r}
imp.temp <- abs(rf$importance)
t <- order(imp.temp, decreasing = TRUE)

plot(c(1:nrow(ex2)), imp.temp[t], log="x", cex.main = 1.5, 
     xlab = "gene rank", ylab="variable importance", cex.lab = 1.5,
     phc = 16, main = "HP vs SP subset results")
```

Accuracy race between RF, LASSO and LDA classification

```{r}
f <- factor(diet_vs_diet_01)
tt40 <- rowttests(ex3, f)
keepers <- which(tt40$p.value<0.1)
tex = t(ex3)
dat <- cbind(as.data.frame(tex), f)

colnames(dat)[ncol(dat)] <- "CONDITION"
n.HP <- 10
n.SP <- 12

control <- trainControl(method="cv", number = 10)
metric <- "Accuracy"

fit.lda <- caret::train(CONDITION~., data=dat, method="lda", metric=metric, trControl=control)
fit.rf <- caret::train(CONDITION~., data=dat, method="rf", metric=metric, trControl=control)
fit.lasso <- caret::train(CONDITION~., 
                          data=dat, 
                          method="glmnet", 
                          family="binomial", 
                          metric=metric, 
                          trControl=control, 
                          tuneGrid = expand.grid(alpha = 1, lambda = seq(0, 1, by=0.05)))
fit.ridge <- caret::train(CONDITION~. ,
                          data=dat, 
                          method = 'glmnet', 
                          faimily = "binomial", 
                          metric = metric, 
                          trControl=control,
                          tuneGrid = expand.grid(alpha = 0, lambda =  seq(0, 1, by=0.05))) 


results <- resamples(list(LDA=fit.lda, RF=fit.rf, LASSO=fit.lasso, RIDGE = fit.ridge))
summary(results)
ggplot(results) + labs(y="Accuracy")

```
testing regression lambda parameters

```{r}
test <- sample(1:(n.HP + n.SP), 5)
train <- setdiff(1:(n.HP + n.SP), test)

# plot of the coefficients for different values of lambda
fit = glmnet(dat[train,], diet_vs_diet_01[train], alpha = 0.7,  standardize = F, family = "binomial")
plot(fit, xvar = "lambda", label = T)

# this runs inside the cross validation
cfit = cv.glmnet(data.matrix(dat[train,]), alpha = 0.7, diet_vs_diet_01[train], standardize = F, family = "binomial")
plot(cfit)

# occasionally you find params with beta != zero
# coef(cfit, s=cfit$lambda.min)

predict(fit, data.matrix(dat[test,]), type = "class", s=cfit$lambda.min)

pred2 <- predict(fit, data.matrix(dat[test,]), type = "response", s = cfit$lambda.min)
plot(performance(prediction(pred2, diet_vs_diet_01[test]), "tpr", "fpr"))
auc.tmp <- performance(prediction(pred2, diet_vs_diet_01[test]), "auc")
auc <- as.numeric(auc.tmp@y.values)
```

# rScudo

```{r}
n.top.bottom = 25
N_ = 0.4

set.seed(123)
inTrain <- createDataPartition(f, list = F)
trainData <- ex3_unique_probes[, inTrain]
testData <- ex3_unique_probes[, -inTrain]

trainRes <- scudoTrain(trainData, groups = f[inTrain], 
                       nTop = n.top.bottom, nBottom = n.top.bottom, alpha = 0.05)

upSignatures(trainRes)[1:5, 1:5]
consensusUpSignatures(trainRes)[1:5, ]

# training
trainNet <- scudoNetwork(trainRes, N= N_)
scudoPlot(trainNet, vertex.label = NA)

# validation with test 
testRes <- scudoTest(trainRes, testData, f[-inTrain])
                     # ntop = 25, nbottom = 25)
testNet <- scudoNetwork(testRes, N = N_)
scudoPlot(testNet, vertex.label = NA)

# indentify clusters on map
testClust <- igraph::cluster_spinglass(testNet, spins = 2)
plot(testClust, testNet, vertex.label = NA)

# perform classification
classRes <- scudoClassify(trainData, testData, N= N_,
                          nTop = n.top.bottom, nBottom = n.top.bottom,
                          trainGroups = f[inTrain], alpha = 0.5)
caret::confusionMatrix(classRes$predicted, f[-inTrain])

```

## rScudo + caret

```{r}

model <- scudoModel(nTop = (2:6)*10, nBottom = (2:6)*10, N = N_)
control <- caret::trainControl(method = "cv", number = 5, summaryFunction = caret::multiClassSummary)
cvRes <- caret::train(x=t(trainData), y=f[inTrain], method = model, trControl = control)

testRes <- scudoTest(trainRes, testData, f[-inTrain],
cvRes$bestTune$nTop,
cvRes$bestTune$nBottom5)
testNet <- scudoNetwork(testRes, N = 0.5)
scudoPlot(testNet, vertex.label = NA)

classRes <- scudoClassify(ex3_unique_probes[, inTrain], ex3_unique_probes[,-inTrain], N_,
                          cvRes$bestTune$nTop, cvRes$bestTune$nBottom,
                          f[inTrain], alpha = 0.05)
caret::confusionMatrix(classRes$predicted, f[-inTrain])
                               
```

preparing lists of genes selected for each method that has been used till now

```{r}
# include in this list the first 200 genes that are considered important for the random forest classification

# rf important genes for classification
rf_importance <- varImp(fit.rf, scale=FALSE)
t <- order(rf_importance$importance$Overall, decreasing = TRUE)
rf_importance_list = c()
for (i in 1:200){
  rf_importance_list <- append(rf_importance_list, 
                                  strsplit(rownames(rf_importance$importance)[t[i]], "`")[[1]][2])
}

# lda important genes for classification
lda_importance <- varImp(fit.lda, scale=FALSE)
t <- order(lda_importance$importance$X0, decreasing = TRUE)
lda_importance_list <- rownames(lda_importance$importance[t[1:200], ])

# lasso important genes for classification
lasso_importance <- varImp(fit.lasso, scale=FALSE)
t <- order(lasso_importance$importance$Overall, decreasing = TRUE)
# these motherfuckers using different notations i hate it
lasso_importance_list <- c()
for (i in 1:200){
  lasso_importance_list <- append(lasso_importance_list, 
                                  strsplit(rownames(lasso_importance$importance)[t[i]], "`")[[1]][2])
}


# ridge important genes for classification
ridge_importance <- varImp(fit.ridge, scale=FALSE)
t <- order(ridge_importance$importance$Overall, decreasing = TRUE)
# these motherfuckers using different notations i hate it
ridge_importance_list <- c()
for (i in 1:200){
  ridge_importance_list <- append(ridge_importance_list, 
                                  strsplit(rownames(ridge_importance$importance)[t[i]], "`")[[1]][2])
}

# lapply(ridge_importance_list, write, "C:/Users/vitto/Documents/uni/IV/second semester/Network based DA/project_NBDA/ridge_list.txt", append=TRUE, ncolumns=1000)
# lapply(lda_importance_list, write, "C:/Users/vitto/Documents/uni/IV/second semester/Network based DA/project_NBDA/lda_list.txt", append=TRUE, ncolumns=1000)
# lapply(rf_importance_list, write, "C:/Users/vitto/Documents/uni/IV/second semester/Network based DA/project_NBDA/rf_list.txt", append=TRUE, ncolumns=1000)
# missing for SCUDO but performs like shit so whatever ^\/(`v`)\/^

```


# Network analysis


```{r}
# load ensembl data
ensembl <- useMart("ensembl")
ensembl <- useEnsembl(biomart = "ensembl", 
                      dataset = "hsapiens_gene_ensembl", 
                      mirror = "useast")
# look for conversion table for probes
ensembl_convertion_table <- getBM(attributes = c('affy_hugene_1_0_st_v1', 'hgnc_symbol'),
      filters = 'affy_hugene_1_0_st_v1', 
      # values = rf_importance_list,  
      values = ridge_importance_list, 
      mart = ensembl)

# remove empty labels
ensembl_convertion_table <- ensembl_convertion_table[!(ensembl_convertion_table$hgnc_symbol == ""), ]

# insert p-value column
ensembl_convertion_table["p_value"] <- NA

# for each selected gene, calculate p-value
# for (target in rf_importance_list){
for (target in ridge_importance_list){
  HP = c()
  NP = c()
  for (j in ER_positions){
    if (group_labels[j] == "ER_HP") {
      HP <- append(HP, 2^ex[target, j])
    } else if (group_labels[j] == "ER_SP"){
      NP <- append(NP, 2^ex[target, j])
    }
  }
  t_NP_HP <- t.test(HP, NP)
  
  # insert p-value in correct row
  for (i in 1:nrow(ensembl_convertion_table)){
    # print(instance)
    if (ensembl_convertion_table[i, ]$affy_hugene_1_0_st_v1 == target){
      ensembl_convertion_table[i, ]$p_value = t_NP_HP$p.value
    }
  }
}

# run_pathfindR
pathfindR_results <- run_pathfindR(ensembl_convertion_table[, c("hgnc_symbol", "p_value")], 
                                   iterations = 1)
pathfindR_cluster <- cluster_enriched_terms(pathfindR_results)
term_gene_graph(pathfindR_results)
```
```{r}
visualize_terms(pathfindR_results, hsa_KEGG = F)

```

```{r}
lapply(ensembl_convertion_table$hgnc_symbol, write, "C:/Users/vitto/Documents/uni/IV/second semester/Network based DA/project_NBDA/nchg_names.txt", append=TRUE, ncolumns=1000)

```


